---
title: "Advanced usage of `scp`"
author:
    - name: Laurent Gatto
    - name: Christophe Vanderaa
output:
    BiocStyle::html_document:
        self_contained: yes
        toc: true
        toc_float: true
        toc_depth: 2
        code_folding: show
bibliography: scp.bib
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('scp')`"
vignette: >
    %\VignetteIndexEntry{Advanced usage of `scp`}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL 
    ## cf https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```

# About this vignette

This vignette is dedicated to advanced users and to method developers.
It assumes that you are already familiar with `QFeatures` and `scp` 
and that you are looking for more flexibility in the analysis of your 
single-cell proteomics (SCP) data. In fact, `scp` provides wrapper
functions around generic functions and metrics. However, advanced 
users may want to apply or develop their own features. The `QFeatures`
class offers a flexible data container while guaranteeing data
consistency. 

In this vignette, you will learn how to: 

- Modify the quantitative data
- Modify the sample annotations
- Modify the feature annotations
- Create a new function for `scp` 

As a general guideline, you can add/remove/update data in a 
`QFeatures` in 4 main steps: 

1. Gather the data to change and other required data involved in the 
   processing.
2. Apply the transformation/computation.
3. Insert the changes in the `QFeatures` object. 
4. Make sure the updated `QFeatures` object is still valid. 

To illustrate the different topics, we will load the `scp1` example 
data.

```{r, message = FALSE}
library(scp)
data("scp1")
scp1
```

# Modify the quantitative data

To illustrate how to modify quantitative data, we will implement a
`normByType` function that will normalize the feature (row) for each
cell type separately. This function is probably not relevant for a 
real case analysis, but it provides a good example of a custom data 
processing. The process presented in this section is applicable to any
custom function that takes at least a **matrix-like object** as input 
and returns a matrix-like object as output. 

```{r}
normByType <- function(x, type) {
    ## Check argument
    stopifnot(length(type) == ncol(x))
    ## Normalize for each type separately
    for (i in unique(type)) {
        ## Get normalization factor
        nf <- rowMedians(x[, type == i], na.rm = TRUE)
        ## Perform normalization
        x[, type == i] <- x[, type == i] / nf
    }
    ## Return normalized data
    x
}
```

Suppose we want to apply the function to the `proteins` assay, we need
to first extract that assay. We here need to transfer the sample 
annotations from the `QFeatures` object to the extracted 
`SingleCellExperiment` in order to get the sample types required by
the `normByType` function. We therefore use `getWithColData`. 

```{r}
sce <- getWithColData(scp1, "proteins")
sce
```

Next, we can apply the data transformation to the quantitative data. 
As mentioned above, our function expects a matrix-like object as an 
input, so we use the `assay` function. We then update the 
`SingleCellExperiment` object. 

```{r}
mnorm <- normByType(assay(sce), type = sce$SampleType)
assay(sce) <- mnorm
```

We are now faced with 2 possibilities: either we want to create a new 
assay or we want to overwrite an existing assay. In both cases we need
to make sure your data is still valid after data transformation. 

## Create a new assay

Creating a new assay has the advantage that you don't modify an 
existing assay and hence limit the risk of introducing inconsistency
in the data and avoid losing intermediate steps of the data 
processing. 

We add the transformed assay using the `addAssay` function, then link
the parent assay to the transformed assay using `addAssayLinkOneToOne`.
Note that if each row name in the parent assay does not match exactly 
one row in the child assay, you can also use `addAssayLink` that will 
require a linking variable in the `rowData`. 

```{r}
scp1 <- addAssay(scp1, sce, name = "proteinsNorm")
scp1 <- addAssayLinkOneToOne(scp1, from = "proteins", to = "proteinsNorm")
scp1
```

## Overwrite an existing assay

Overwriting an existing assay has the advantage to limit the memory 
consumption as compared to adding a new assay. You can overwrite an 
assay simply by replacing the target assay in its corresponding slot.

```{r, eval = FALSE}
scp1[["proteins"]] <- sce
```

## Check for validity 

Applying custom changes to the data increases the risk for data 
inconsistencies. You can however verify that a `QFeatures` object is
still valid after some processing thanks to the `validObject` function.

```{r}
validObject(scp1)
```

If the function detects no issues in the data, it will return `TRUE`.
Otherwise the function will throw an informative error that should 
guide the user to identifying the issue.

# Modify the sample annotations

To illustrate how to modify the sample annotations, we will compute the 
average expression in each sample and include to the `colData` of the 
`QFeatures` object. This is typically performed when computing QC 
metrics for sample filtering. So, we first extract the required data, 
in this case the quantitative values, and compute the sample-wise 
average protein expression.

```{r}
m <- assay(scp1, "proteins")
meanExprs <- colMeans(m, na.rm = TRUE)
meanExprs
```

Next, we insert the computed averages into the `colData`. You need to
make sure to match sample names because an extracted assay may not 
contain all samples and may be in a different order compared to the
`colData`. 

```{r}
colData(scp1)[names(meanExprs), "meanProtExprs"] <- meanExprs
```

The new sample variable `meanProtExprs` is now accessible for filtering
or plotting. The `$` operator makes it straightforward to access the
new data. 

```{r}
hist(log2(scp1$meanProtExprs))
```

To make sure that the process did not corrupt the `colData`, we advise
to verify the data is still valid. 

```{r}
validObject(scp1)
```

# Modify the feature annotations

We will again illustrate how to modify the feature annotations with an 
example. We here demonstrate how to add the number of samples in which
each feature is detected for the three first assays (PSM assays). The
challenge here is that the metric needs to be computed for each assay
separately and inserted in the corresponding assay. 

We will take advantage of the replacement function for `rowData` as 
implemented in `QFeatures`. It expects a list-like object where names 
indicate in which assays we want to modify the `rowData` and each 
element contains a table with the replacement values. 

We therefore compute the metrics for each assay separately and store
the results in a named `List`. 

```{r}
## Initialize the List object that will store the computed values
res <- List()
## We compute the metric for the first 3 assays
for (i in 1:3) {
    ## We get the quantitative values for the current assay
    m <- assay(scp1[[i]])
    ## We compute the number of samples in which each features is detected
    n <- rowSums(!is.na(m) & m != 0)
    ## We store the result as a DataFrame in the List
    res[[i]] <- DataFrame(nbSamples = n)
}
names(res) <- names(scp1)[1:3]
res
res[[1]]
```

Now that we have a `List` of `DataFrame`s, we can update the object.

```{r}
rowData(scp1) <- res
```

The new feature variable `nbSamples` is now accessible for filtering
or plotting. The `rbindRowData` function facilitates the access the
new data. 

```{r}
library("ggplot2")
rd <- rbindRowData(scp1, i = 1:3)
ggplot(data.frame(rd)) +
    aes(x = nbSamples) +
    geom_histogram(bins = 16) +
    facet_wrap(~ assay)
```

To make sure that the process did not corrupt the `rowData` in any 
assay, we advise to verify the data is still valid. 

```{r}
validObject(scp1)
```

# Create a new function for `scp` 

The modifying data in a `QFeatures` involves a multiple-step process.
Creating a wrapper function that would take care of those different 
steps in a single line of code is a good habit to reduce the length of
analysis scripts and hence making it easier to understand and less 
error-prone. 

We will wrap the last example in a new function that we call
`computeNbDetectedSamples`. 

```{r}
computeNbDetectedSamples <- function(object, i) {
    res <- List()
    for (ii in i) {
        m <- assay(object[[ii]])
        n <- rowSums(!is.na(m) & m != 0)
        res[[ii]] <- DataFrame(nbSamples = n)
    }
    names(res) <- names(object)[i]
    rowData(object) <- res
    stopifnot(validObject(object))
    object
}
```

Thanks to this new function, the previous section now simply boils 
down to running: 

```{r}
scp1 <- computeNbDetectedSamples(scp1, i = 1:3)
```

Keep in mind a few recommendations when creating a new function for 
`scp`: 

- The function should take a `QFeatures` object as input (and more 
  arguments if needed) and return a `QFeatures` object as output. This
  will make workflows much easier to understand.
- Allow user to select assays (if required) either as numeric, character,
  or logical. 
- Use conventional argument names: when naming an argument, try to 
  match the names that already exist. For instance, selecting assays 
  is passed through the `i` argument, selecting `rowData` variables is 
  passed through `rowvars` and selecting `colData` variables is passed
  through `colvars`.
- Follow the `rformassspectrometry` 
  [coding style](https://rformassspectrometry.github.io/RforMassSpectrometry/articles/RforMassSpectrometry.html#coding-style)

# What's next?

So you developed a new metric or method and believe it might benefit 
the community? We would love to hear about your improvements and 
eventually include your new functionality into `scp` or associate your
new package to our documentation. Please, raise an
[issue](https://github.com/UCLouvain-CBIO/scp/issues/new/choose) in
our Github repository to suggest your improvements or, better, submit 
your code as a 
[pull request](https://github.com/UCLouvain-CBIO/scp/compare). 

# Session information {-}

```{r setup2, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "",
    crop = NULL
)
```


```{r sessioninfo, echo=FALSE}
sessionInfo()
```

# License

This vignette is distributed under a 
[CC BY-SA license](https://creativecommons.org/licenses/by-sa/2.0/) 
license.

# Reference
