---
title: "Reporting missing values for Single Cell Proteomics"
author:
    - name: Christophe Vanderaa
    - name: Laurent Gatto
output:
    BiocStyle::html_document:
        self_contained: yes
        toc: true
        toc_float: true
        toc_depth: 2
        code_folding: show
bibliography: scp.bib
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('scp')`"
vignette: >
    %\VignetteIndexEntry{Reporting missing values}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL
)
```

# Introduction

This vignette demonstrates how to use `scp` to report missing values,
following our recommendations in @Vanderaa2023-gu. Briefly, we 
recommend reporting at least 4 metrics: 

- Total sensitivity, i.e. the total number of features found in the 
dataset
- Local sensitivity, i.e. the number of features per cell
- Data completeness, i.e. the proportion of values that are not 
missing
- Number of samples

We will also demonstrate how to estimate total sensitivity when the 
number of samples is too low and how to report data consistency using
the distribution of Jaccard indices.

In this vignette, we will assume you are familiar with the `scp` 
framework. If this is not the case, we suggest you first read the 
[`introduction vignette`](https://uclouvain-cbio.github.io/scp/articles/scp.html).

# Minimal data processing

First, we load the `scp` package and retrieve a real-life dataset from
the `scpdata` package.

```{r}
library("scp")
library("scpdata")
leduc <- leduc2022()
```

Next, we reduce the size of the dataset to the 30 first acquisitions.
This allows for a fast execution of the code while still being a
representative demonstration on a real dataset. We also keep only the
feature annotations that will be used later in the vignette.

```{r}
leduc <- leduc[, , 1:30]
leduc <- selectRowData(leduc, c(
    "Sequence", "Leading.razor.protein", "Reverse", 
    "Potential.contaminant", "PEP"
))
```

This is the actual minimal processing: filtering contaminant and 
low-quality features, replacing zeros by missing values, we keep only
samples that correspond to single cells and we remove the feature 
absent in all samples.

```{r}
leduc <- filterFeatures(leduc, ~ Reverse != "+" &
                           Potential.contaminant != "+" &
                           PEP < 0.01)
leduc <- zeroIsNA(leduc, i = names(leduc))
leduc <- filterNA(leduc, i = names(leduc), pNA = 0.9999)
leduc <- filterNA(leduc, i = names(leduc), pNA = 0.9999)
leduc <- subsetByColData(
    leduc, leduc$SampleType %in% c("Monocyte", "Melanoma cell")
)
leduc <- dropEmptyAssays(leduc)
```

# Peptide identification data

Next, we build the peptide identification matrix, that is a matrix
with single cells as columns and peptides as rows. We first need data 
at peptide level, hence we aggregate the peptide to spectrum match 
data to peptide data. We use a counting functions for aggregation, 
meaning that we generate spectral count data for peptides. Then, we
join all runs in a single assay. 

```{r}
leduc <- aggregateFeatures(
    leduc, i = names(leduc), 
    name = paste0("spectral_counts_", names(leduc)),
    fcol = "Sequence", 
    fun = MsCoreUtils::colCounts
)
leduc <- joinAssays(
    leduc, i = grep("^spec", names(leduc)), 
    name = "spectral_counts"
)
```

Once all assay are joined, we extract that assay and replace any 
observed count by 1 and any missing value by 0. This is the peptide
identification matrix.

```{r}
ids <- getWithColData(leduc, "spectral_counts")
assay(ids) <- ifelse(is.na(assay(ids)), 0, 1)
```

# Report missing values

We can now measure the for metrics. We recommend computing these for
each cell type separately, since biological properties specific to the
cell type could influence the outcome.

```{r}
(metrics <- sapply(unique(ids$SampleType), function(st) {
    x <- assay(ids)[, ids$SampleType == st]
    c(
        LocalSensitivityMean = mean(colSums(x != 0)),
        LocalSensitivitySd = sd(colSums(x != 0)),
        TotalSensitivity = sum(rowSums(x) != 0),
        Completeness = mean(x != 0),
        NumberCells = ncol(x)
    )
}))
```

**What about proteins?** To generate the same table for proteins, we 
need to rebuild an identification matrix for peptides. In this case, 
you simply need to change the aggregation call

```{r, eval=FALSE}
leduc <- aggregateFeatures(
    leduc, i = names(leduc), 
    name = paste0("spectral_counts_", names(leduc)),
    fcol = "Sequence", 
    fun = MsCoreUtils::colCounts
)
```

to

```{r, eval=FALSE}
leduc <- aggregateFeatures(
    leduc, i = names(leduc), 
    name = paste0("spectral_counts_", names(leduc)),
    fcol = "Leading.razor.protein", 
    fun = MsCoreUtils::colCounts
)
```

All code that follows aggregation is exactly the same.

# Advanced criteria

On top of the four metrics computed in the previous section, we
also recommended exploring the jaccard indices and assessing whether
enough single cells are acquired to accurately estimate the total 
sensitivity. 

## Jaccard index distribution

We first need a custom function that computes the Jaccard index for 
all pairs of columns given an identification matrix. The Jaccard index
in this context is the number of features shared by two columns 
divided by the number of distinct features identified in any of the 
two columns. This provides a good measure of how consistent the 
identifications are across single-cells.

```{r}
jaccardIndexFromIdMatrix <- function(idMatrix) {
    vectorSizes <- colSums(idMatrix)
    pwVectorSizes <- sapply(vectorSizes, function(x) vectorSizes + x)
    unionSize <- crossprod(idMatrix)
    jacc <- unionSize / (pwVectorSizes - unionSize)
    jacc[upper.tri(jacc)] ## jacc is a symmetric matrix
}
```

Next we plot the distribution of the jaccard indices. Again, 
biological differences between cell types may decrease the consistency
between single cells. Therefore, we compute and plot the Jaccard index
for each cell type separately.

```{r}
par(mfrow = c(2, 1))
for (st in unique(ids$SampleType)) {
    x <- assay(ids)[, ids$SampleType == st]
    hist(jaccardIndexFromIdMatrix(x), main = st, 
         xlab = "Jaccard index", breaks = 50)
}
```

The Jaccard index is mainly distributed across 50 and 60 \%, meaning
that about half of the features are consistently found across 
single-cells within the same cell type. Note also that some pairs of
cells have perfect consistency. These are pairs of cells from the same
acquisition runs that were multiplexed together with TMT labelling.

## Assessing the total sensitivity

To assessing whether we can accurately estimate the total sensitivity,
we generate a cumulative sensitivity curve (CSC). The functions below
allows to generate that curve provided an identification matrix. The
algorithm samples columns from the matrix in gradually increasing 
numbers (determined by `nsteps`). Each sampling is repeated `niters` 
times to account for the stochasticity of the approach. 

```{r}
getSteps <- function(assays, nsteps) {
    maxn <- length(assays)
    steps <- seq(1, maxn, length.out = min(maxn, nsteps))
    round(steps)
}
groupSamples <- function(x, by) {
    sapply(unique(by), function(i) {
        ifelse(rowSums(x[, by == i]) == 0, 0, 1)
    })
}
cumulativeSensitivity <- function(idMatrix, niters, nsteps, 
                                  groupSamplesBy = NULL) {
    out <- list()
    if (!is.null(groupSamplesBy)) {
        idMatrix <- groupSamples(idMatrix, by = groupSamplesBy)
    }
    for (n in getSteps(colnames(idMatrix), nsteps)) {
        for (i in 1:niters) {
            sel <- sample(colnames(idMatrix), n)
            s <- sum(rowSums(idMatrix[, sel, drop = FALSE]) != 0)
            out <- c(
                out,
                list(data.frame(
                    i = i,
                    NumberSamples = n,
                    Sensitivity = s
                ))
            )
        }
    }
    do.call(rbind, out)
}
```

Again, we explore the csc for each cell type separately. Note that we
group the cells by acquisition run (`Set`).

```{r}
csc <- lapply(unique(ids$SampleType), function(st) {
    x <- ids[, ids$SampleType == st]
    out <- cumulativeSensitivity(
        assay(x), niters = 10, nsteps = 30, groupSamplesBy = x$Set
    )
    out$SampleType <- st
    out
})
csc <- do.call(rbind, csc)
```

Next, we plot the cumlulative sensitivity for each cell type 
separately.

```{r}
library("ggplot2")
(plCumS <- ggplot(csc) +
    aes(x = NumberSamples, y = Sensitivity, colour = SampleType) +
    geom_point(size = 1) +
    labs(y = "Sensitivity", x = "Number of MS acquisitions"
    ) +
    theme_minimal() +
    theme(
        legend.position = "top",
        axis.title.x = element_text(vjust = 0)
    ))
```

The cumulative sensitivity does not plateau, although it almost 
reaches it. This means that we slightly underestimate the total 
sensitivity. This is because we have subset the data at the beginning
of the script, and including more datasets allows to reach the 
plateau.

We will now predict the total sensitivity. Since both cell types 
exhibit the same trends, we will perform the prediction for the two 
cell types combined. We first fit an asymptotic regression model to
assess the relationship between the sensitivity and the number of 
MS runs. Then, we predict the response and plot these preduction on
the previous graph.

```{r}
AsymptoticRegressionModel <- nls(
    formula = Sensitivity ~ SSasymp(NumberSamples, Asym, R0, lrc),
    data = csc,
    weights = csc$NumberSamples^2 ## data is heterorscedastic
)
predictedSensitivity <- data.frame(
    NumberSamples = 1:max(csc$NumberSamples) 
)
predictedSensitivity$Sensitivity <- predict(
    AsymptoticRegressionModel,
    newdata = predictedSensitivity
)
plCumS +
    geom_line(data = predictedSensitivity, colour = "red")
```

We finally compute the predicted total sensitivity, that is 
sensitivity if we had an infinite number of samples.

```{r}
predict(
    AsymptoticRegressionModel,
    newdata = data.frame(NumberSamples = Inf) 
)
```

The total sensitivity is predicted to be 7282 peptides. Compared to 
the total sensitivity of 7111 reported in the previous section, we 
underestimate the total sensitivity by about 150 peptides.

# Citation

```{r citation}
citation("scp")
```

# License

This vignette is distributed under a 
[CC BY-SA license](https://creativecommons.org/licenses/by-sa/2.0/) 
license.

# Reference
